---
title: "ST558_Project1"
author: "Stephanie Shuai"
format: html
editor: visual
---

# Introduction

This project aims to use the U.S. Census API to extract and analyze individualized data from the U.S. Census PUMS dataset. The goal is to build a flexible data processing workflow that enables researches to query, process, and analyze survey data using weighted statistical methods. Due to the impact of the COVID-19 pandemic, data from 2020 has been excluded, providing an opportunity to compare data patterns before and after the pandemic.

# Load Libraries

```{r}
library(httr)
library(jsonlite)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Data Processing

We begin by constructing a data processing pipeline that includes data retrieval, parsing, and cleaning.

```{r}
url <- "https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,AGEP,PWGTP,SCHL&key=1c79cd9423b45d9bcae0f21f1c8080b2d8560a24"

# ---- GET Function ----
get_census_data <- function(url){
  res <- httr::GET(url)
  if(httr::status_code(res) != 200){
    stop("API request failed")
  }
  
  return(res)
}

# ---- Parse Function ----
parse_to_tibble <- function(res) {
  txt <- content(res, as = "text", encoding = "UTF-8")
  json <- fromJSON(txt)
  
  df <- as.data.frame(json)
  colnames(df) <- df[1,]
  df <- df[-1,]
  
  as_tibble(df)
}

process_geo_subset <- function(geo = "state", geo_filter = "*") {
  if(geo_filter == "all") geo_filter <- "*"
  paste0("for=", geo, ":", geo_filter)
}
```

# Main Query Function

```{r}
query_census <- function(
    year = 2022, 
    num_var = c("AGEP","PWGTP"),
    cat_var = "SEX",
    geography = "state",
    geo_filter = "*"
  ){
  
  # ---- Validation ----
  if (year < 2010 | year > 2022 | year == 2020 ){
    stop("Invalid year")
  }
  
  num <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP","JWMNP", "PWGTP")
  cat <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  if (!all(num_var %in% num)) stop("Invalid numeric var")
  if (!all(cat_var %in% cat)) stop("Invalid categorical var")
  
  if (!"PWGTP" %in% num_var ) {
    num_var <- c(num_var, "PWGTP")
  }
  
  geo <- c("state", "region", "division")
  if (!geography %in% geo) {
    stop("Invalid geography")
  }
  
  vars <- na.omit(unique(c(num_var, cat_var)))
  var_string <- paste(vars, collapse = ",")

  geo_string <- paste0("for=", geography, ":", geo_filter)
  
  url <- paste0(
    "http://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    var_string,
    "&",
    geo_string,
    "&key=1c79cd9423b45d9bcae0f21f1c8080b2d8560a24"
    )
  res <- get_census_data(url)
  df <- parse_to_tibble(res)
  
  df$AGEP <- as.numeric(df$AGEP)
  df$PWGTP <- as.numeric(df$PWGTP)
  df$SEX <- factor(df$SEX, levels = c("1","2"), labels = c("Male","Female"))

  for (v in num_var){
    df[[v]] <- as.numeric(df[[v]])
  }
  for (v in cat_var){
    df[[v]] <- as.factor(df[[v]])
  }
  class(df) <- c("census", class(df))
  
  return(df)
}

# ---- Multi-Year Function ----
query_census_multi <- function(years, ...){
  if (any(years == 2020)) stop("Exclude 2020")
  if (any(years < 2010 | years > 2022)) stop("Invalid year")
  
  all_data <- lapply(years, function(y){
    df <- query_census(year = y, ...)
    df$year <- y
    df
  })
  dplyr::bind_rows(all_data)
}
```

## Writing a Generic Function for Summarzing

```{r}
# ---- Summary Function ----
summary.census <- function(object, num_var , cat_var ){
  x <- object[[num_var]]
  w <- object$PWGTP
  
  w_mean <- sum(x*w)/sum(w)
  cat_counts <- table(object[[cat_var]])
  
  list(
    weighted_mean = sum(x*w) /sum(w),
    weighted_sd = sqrt(sum(w* x^2) / sum(w) - w_mean^2),
    category_counts = cat_counts
  )
}
# ---- Plot Function ----
plot.census <- function(x, num_var, cat_var) {
  if(nrow(x) > 5000){
    x <- x %>% slice_sample(n=5000)}
  
  ggplot(x, aes(x = .data[[cat_var]], 
                y = .data[[num_var]], 
                weight = PWGTP)) +
    geom_boxplot() + 
    labs(x = cat_var, y = num_var) + 
    theme_minimal()
}
```

## Example

```{r}
df <- query_census(
  num_var = c("AGEP","PWGTP"),
  cat_var = c("SEX", "SCHL"))

head(df)

summary(df, num_var = "AGEP", cat_var = "SEX")

plot(df, num_var = "AGEP", cat_var = "SEX")

# Multi-year example
df_multi <- query_census_multi(
  years = c(2019, 2021),
  num_var = c("AGEP", "PWGTP"),
  cat_var = "SEX",
  geography = "state",
  geo_filter = "37"
)

head(df_multi)
```

# Narrative

Excluding data from 2020 helps facilitate a comparison of demographic variables before and after the COVID-19 pandemic. The results show that the age distribution remained stable across all years, indicating consistency in the population structure.

In contrast, the GRPIP index showed a steady downward trend from 2019 to 2022, suggesting changes in housing affordability or income dynamics. However, the high level of variability indicates that the impact varied among individuals.

The gender distribution remained stable, and enrollment patterns showed little change. However, education attainment shifted toward higher levels.

Overall, the findings suggest that demographic variables remain stable, while socioeconomic indicators are more sensitive to external shocks such as the COVID-19 pandemic.
